---
title: "tidyverse_intro"
output: html_document
---

This is a setup for when you need to _Knit_ files - into PDF, HTML, etc.
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


The following chunk of code is for installing required packages.

```{r load_packages}
# A slightly more elegant way of loading packages than what I showed during workshop.

# It's a function that checks for you as to whether packages need to be installed.
# You don't need to know how it works to use it. But explanation is in comments.

check_packages <- function(pkg) {
  # This vector checks packages that are not installed yet.
  new_pkg <- pkg[!(pkg %in% installed.packages()[, "Package"])]

  # If there are new packages then it installs them along with dependencies.
  if (length(new_pkg)) {
    install.packages(new_pkg, dependencies = TRUE)
  }
  # Apply statement applies each package name to require, so require('tidyverse')
  # And now they're loaded into your gloabal environment
  sapply(pkg, require, character.only = TRUE)
}

# Here inside this vector you load all the packages you need.
packages <- c("tidyverse", "zoo", "hrbrthemes")

# Then you run the check_packages function and it loads everything.
check_packages(packages)
```

Now, we have the packages we need. Next, we start reading the data. All the data downloaded from *Inforshare* is stored inside the `raw_data` folder. These files are untouched.

We started by figuring out what was wrong with the data.

Then, how we might structure it.

The first step is to have clean column names that we can reuse later.

```{r column_names}
# Firstly, we want to prepare column names - in this case country names.

# Extract country names, to use as column names for data frames
nz_arrival_names <- read_csv("raw_data/nz_2000_arrivals.csv", col_names = FALSE, n_max = 1, skip = 2)

# Convert to character vector
country_names <- as.character(as.vector(nz_arrival_names))

# Remove the data frame from the global environment.
rm(nz_arrival_names)

# Assign first value to the character vector for the date row
country_names[1] <- "month"

# You can check this code one step at a time by yourself.
```

Now we're going to walk step by step through how we can clean the data set.

First we read the CSV file, skipping first four rows and stoping at 206.

```{r}
nz_arrivals_2000 <- read_csv("raw_data/nz_2000_arrivals.csv", col_names = FALSE, n_max = 206, skip = 4)
```

Next we assign the column names we created in `country_names` to the data frame we just read.

```{r}
names(nz_arrivals_2000) <- country_names


# We can see what the dataset looks like -
nz_arrivals_2000
```


Then we rename the column names that are inconsistent. `rename` comes from `dplyr` package.


We're also piping the dataframe. One of the ways to think about this is, `data_frame` (then) `some_function` (then) `another_function` - whereby output of each step becomes the input for the next

```{r}

nz_arrivals_2000 %>% rename(
  Total = `TOTAL ALL COUNTRIES OF RESIDENCE`,
  China = `China, People's Republic of`,
  Oceania = `OCEANIA`,
  Asia = `ASIA`,
  Europe = EUROPE,
  `Africa and the Middle East` = `AFRICA AND THE MIDDLE EAST`,
  `Hong Kong` = `Hong Kong (Special Administrative Region)`,
  `German Democractic Republic` = `Germany, Democratic Republic of`,
  Americas = AMERICAS,
  `South Korea` = `Korea, Republic of`,
  Palestine = `Gaza Strip/Palestine/West Bank`,
  `Not Stated` = `NOT STATED`
)





  mutate(month = gsub("M", "/", month, fixed = TRUE)) %>%
  separate(month, c("year", "mon"), sep = "/") %>%
  mutate(day = "01") %>%
  unite(month, day, mon, year, sep = "/") %>%
  mutate(month = as.Date(month, format = "%d/%m/%Y")) %>%
  mutate(count = as.integer(count)) %>%
  filter(country == "Australia") %>%
  ggplot(., aes(x = month, y = count, color = "red")) + geom_line() + theme_ipsum() + theme(legend.position = "none")
```


Next we gather the variables, into country and counts column. Going from a wide to long format.

To do this we use `gather` function from `tidyr` package. For opposite action, long to wide format, we'd use `spread` function.

```{r}
nz_arrivals_2000 %>%
  rename(
    Total = `TOTAL ALL COUNTRIES OF RESIDENCE`,
    China = `China, People's Republic of`,
    Oceania = `OCEANIA`,
    Asia = `ASIA`,
    Europe = EUROPE,
    `Africa and the Middle East` = `AFRICA AND THE MIDDLE EAST`,
    `Hong Kong` = `Hong Kong (Special Administrative Region)`,
    `German Democractic Republic` = `Germany, Democratic Republic of`,
    Americas = AMERICAS,
    `South Korea` = `Korea, Republic of`,
    Palestine = `Gaza Strip/Palestine/West Bank`,
    `Not Stated` = `NOT STATED`
  ) %>%
  gather(country, count, -month)
```

At this stage, our data set is in a tidy format.

However, data types are inconsistent.

Dates and integers are not properly formatted.

Here we do series of steps to format the date properly. We use `gsub` to replace `M` with `\` - then we go through steps to get a date format. There are multiple ways of approaching this step. You can probably come up with a better one and reduce the number of steps.

```{r}
nz_arrivals_2000 %>%
  rename(
    Total = `TOTAL ALL COUNTRIES OF RESIDENCE`,
    China = `China, People's Republic of`,
    Oceania = `OCEANIA`,
    Asia = `ASIA`,
    Europe = EUROPE,
    `Africa and the Middle East` = `AFRICA AND THE MIDDLE EAST`,
    `Hong Kong` = `Hong Kong (Special Administrative Region)`,
    `German Democractic Republic` = `Germany, Democratic Republic of`,
    Americas = AMERICAS,
    `South Korea` = `Korea, Republic of`,
    Palestine = `Gaza Strip/Palestine/West Bank`,
    `Not Stated` = `NOT STATED`
  ) %>%
  gather(country, count, -month) %>% 
  mutate(month = gsub("M", "/", month, fixed = TRUE)) %>%
  separate(month, c("year", "mon"), sep = "/") %>%
  mutate(day = "01") %>%
  unite(month, day, mon, year, sep = "/") %>%
  mutate(month = as.Date(month, format ="%d/%m/%Y")) %>%
  mutate(count = as.integer(count))
```


The above code gives us a tidy data format with proper data types.

We can quickly see the patterns by making a chart without necessarily saving anything into a variable.

For example, if we wanted to see arrivals from Australia for NZ citizens since 2000.

We can `filter` and then make a plot using `ggplot`.

```{r}
nz_arrivals_2000 %>%
  rename(
    Total = `TOTAL ALL COUNTRIES OF RESIDENCE`,
    China = `China, People's Republic of`,
    Oceania = `OCEANIA`,
    Asia = `ASIA`,
    Europe = EUROPE,
    `Africa and the Middle East` = `AFRICA AND THE MIDDLE EAST`,
    `Hong Kong` = `Hong Kong (Special Administrative Region)`,
    `German Democractic Republic` = `Germany, Democratic Republic of`,
    Americas = AMERICAS,
    `South Korea` = `Korea, Republic of`,
    Palestine = `Gaza Strip/Palestine/West Bank`,
    `Not Stated` = `NOT STATED`
  ) %>%
  gather(country, count, -month) %>% 
  mutate(month = gsub("M", "/", month, fixed = TRUE)) %>%
  separate(month, c("year", "mon"), sep = "/") %>%
  mutate(day = "01") %>%
  unite(month, day, mon, year, sep = "/") %>%
  mutate(month = as.Date(month, format ="%d/%m/%Y")) %>%
  mutate(count = as.integer(count)) %>% 
  filter(country == "Australia") %>% 
  ggplot(., aes(x=month, y=count)) + geom_line()
```

In this plot, we can see a lot of people arrive home at the same time. 

Yep, that's people coming home for Christmas/New Year holidays.

Recently, however these arrivals have gone up - probably signaling a change in patterns. This is a simple way to explore the data set.


Now we have created some code to clean a `data frame` - but we want to reuse this code. We do that by creating a function. We basically generalise the solution to same problem.


```{r data_cleaning_function, message=FALSE, warning=FALSE, include=FALSE}

tidyToLongFormat <- function(input_frame, citizen, travel) {
  names(input_frame) <- country_names

  processed_data <- input_frame %>%
    rename(
      Total = `TOTAL ALL COUNTRIES OF RESIDENCE`,
      China = `China, People's Republic of`,
      Oceania = `OCEANIA`,
      Asia = `ASIA`,
      Europe = EUROPE,
      `Africa and the Middle East` = `AFRICA AND THE MIDDLE EAST`,
      `Hong Kong` = `Hong Kong (Special Administrative Region)`,
      `German Democractic Republic` = `Germany, Democratic Republic of`,
      Americas = AMERICAS,
      `South Korea` = `Korea, Republic of`,
      Palestine = `Gaza Strip/Palestine/West Bank`,
      `Not Stated` = `NOT STATED`
    ) %>%
    gather(country, count, -month) %>%
    mutate(month = gsub("M", "/", month, fixed = TRUE)) %>%
    separate(month, c("year", "mon"), sep = "/") %>%
    mutate(day = "01") %>%
    unite(month, day, mon, year, sep = "/") %>%
    mutate(month = as.Date(month, format = "%d/%m/%Y")) %>%
    mutate(count = as.integer(count)) %>%
    mutate_(
      citizen = !citizen,
      travel = !travel
    )

  processed_data
}
```

